Evaluating the Performance and Energy Efficiency of the COSMO-ART model system

### 2014-04-23 trunk/paper_enahpc14/Introduction.tex
shortens paper length with the deletion of some references

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of   solar   radiation,    thereby   exerting   a   direct   radiative
forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
modification of  cloud properties, cloud fraction  and surface albedo,
causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and    measurement-based    assessments   \citep{Myhre-2009},    large
uncertainties  remain  in   current  estimates  of  aerosol  radiative
forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
Rosenfeld-2013, Sherwood-2013, Stier-2013}.

>>>>

While anthropogenic greenhouse gas emissions are driving unprecedented
major  climate  changes  since   the  mid  20th  Century,  one  factor
overwhelmingly  affects  the  uncertainty  in  defining  human-induced
radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
Although they  are not considered  a heat-trapping greenhouse  gas and
have  shorter  atmospheric lifetimes,  they  significantly modify  the
global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
concentrations impact the climate  system by scattering and absorption
of solar radiation, thereby exerting a direct radiative forcing; or by
the  modification  of cloud  properties,  cloud  fraction and  surface
albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
considerable  progress in  global aerosol  modelling \citep{Mann-2013}
and    measurement-based    assessments   \citep{Myhre-2009},    large
uncertainties  remain  in   current  estimates  of  aerosol  radiative
forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
Rosenfeld-2013, Sherwood-2013, Stier-2013}.

### 2014-04-23 trunk/paper_enahpc14/Sec4:Evaluation.tex
shortens paper length with the deletion of some references

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimisation  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.   The  reason  for  that  is two-fold:  it  has  higher  clock
frequency than  Ivy Bridge (2.6  GHz against 2.2  GHz) and it  aims at
computing speed regardless to  energy consumption. In our experiments,
Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
consumption of Xeon E5 by approximately $7\%$.

>>>>

\subsubsection{Sofware environment}
The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
running on distributed multi-core systems only.  The software stack on
both CSCS  platforms was controlled using the  modules framework which
gives an easy and flexible mechanism  to access to all of the provided
compilers, tools  and applications.  For our  initial benchmarking, we
opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
delivering inferior performance.  In addition, we installed the MPICH2
implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
large  and  complex data  collections.   All  computes  nodes have  an
operating      system      based      on      GNU/Linux      featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
in this study as previous attempts  of its use revealed that it always
led to higher energy-to-solution.\\

@@

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.   The reason for  that is  two-fold: (1)  it has  higher clock
frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
at  computing   speed  regardless   to  energy  consumption.   In  our
experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
reducing the energy consumption of Xeon E5 by approximately $7\%$.
